---
title: "Exploratory Data Analysis of COVID patients Metadata"
author: "Diego Maquieira Vidal (diego.maquieira01@estudiant.upf.edu) and Alberto Domingo Gómez (alberto.domingo01@estudiant.upf.edu)"
date: "2026-03-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This practical focuses on data wrangling, exploratory data analysis, visualization, and critical comparison with published results using real biomedical data.

We will work with the publication [**Proteomic and Metabolomic Characterization of COVID-19 Patient Sera**](https://www.sciencedirect.com/science/article/pii/S0092867420306279)

# Exercise 1  

Download and load Supplementary Table 1 (“Additional Demographical and Baseline Characteristics of COVID-19 Patients and Control Groups”) from the supplementary materials of the publication.
 
## Exercise 1.1  

Reproduce Table 1 from the manuscript using the supplementary data. Match the reported summary statistics   as closely as possible.

Briefly discuss any discrepancies between your results and the published table, and provide possible explanations.

```{r}
# Automated Package Management
# Define the list of required packages
setwd("~/Documents/Workspace/DMI-WorkSpace/assignment-1-dtmk10/")
packages <- c(
  "readxl", "dplyr", "tidyr", "knitr", "kableExtra", 
  "ggplot2", "corrplot", "purrr", "ggpubr", "rstatix", 
  "pheatmap", "grid", "RColorBrewer"
)

suppressMessages(suppressWarnings(invisible(if (!require("readxl")) {install.packages("readxl"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("dplyr")) {install.packages("dplyr"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("tidyr")) {install.packages("tidyr"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("knitr")) {install.packages("knitr"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("kableExtra")) {install.packages("kableExtra"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("ggplot2")) {install.packages("ggplot2"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("corrplot")) {install.packages("corrplot"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("purrr")) {install.packages("purrr"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("ggpubr")) {install.packages("ggpubr"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("rstatix")) {install.packages("rstatix"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("pheatmap")) {install.packages("pheatmap"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("grid")) {install.packages("grid"); require(dendextend)})))
suppressMessages(suppressWarnings(invisible(if (!require("RColorBrewer")) {install.packages("RColorBrewer"); require(dendextend)})))

# Libraries preparation
invisible(library(readxl))
invisible(library(dplyr))
invisible(library(tidyr))
invisible(library(knitr))
invisible(library(kableExtra))
invisible(library(ggplot2))
invisible(library(corrplot))
invisible(library(purrr))
invisible(library(ggpubr))
invisible(library(rstatix))
invisible(library(pheatmap))
```

```{r}
# Load and Preprocess Clinical Metadata
# Define the named vector for all columns to ensure safe importing
st1_types <- c(
  "Patient ID a" = "text", "Proteomics" = "text", "MS ID b" = "text", 
  "MSRep ID c" = "text", "Group d" = "numeric", "COVID-19 patients corhort" = "text", 
  "Metabolomics analysis" = "text", "Metabolomics ID e" = "text", "Onset date f" = "date", 
  "Admission date" = "date", "Date of progression to severe state" = "date", 
  "Sampling date for metabolomics" = "date", "Sampling date for proteomics" = "date", 
  "Discharge Date" = "date", "Sex g" = "numeric", "Age (year)" = "numeric", 
  "BMI h" = "numeric", "Test date...18" = "date", "WBC count, ×109/L" = "numeric", 
  "Lymphocyte count, ×109/L" = "numeric", "Monocyte count, ×109/L" = "numeric", 
  "Platelet count, ×109/L" = "numeric", "Test date...23" = "date", "CRP i, mg/L" = "numeric", 
  "ALT j,  U/L" = "numeric", "AST k, U/L" = "numeric", "GGT l, U/L" = "numeric", 
  "TBIL m, μmol/L" = "numeric", "DBIL n, μmol/L" = "numeric", "Creatinine, μmol/L" = "numeric", 
  "Glucose, mmol/L" = "numeric"
)

# Load the data
suppressWarnings(suppressMessages(df_clinical <- read_excel("data/ts_1_data.xlsx", 
                          sheet = "Clinical_information", 
                          col_types = st1_types)))

# Calculate time intervals and assign descriptive group names
df_clinical <- df_clinical %>%
  mutate(
    # Calculate days. Adding 1 to include both start and end days
    days_onset_to_adm = as.numeric(difftime(`Admission date`, `Onset date f`, units = "days")) + 1,
    days_adm_to_severe = as.numeric(difftime(`Date of progression to severe state`, `Admission date`, units = "days")) + 1,
    
    # Map numeric groups to descriptive factors
    Group_Label = case_when(
      `Group d` == 0 ~ "Healthy Control",
      `Group d` == 1 ~ "Non-COVID-19",
      `Group d` == 2 ~ "Non-Severe COVID-19",
      `Group d` == 3 ~ "Severe COVID-19"
    )
  )

# Summarize Clinical Characteristics
# Helper function to format range safely
get_range <- function(x) {
  if(all(is.na(x))) return(NA_character_)
  paste0(round(min(x, na.rm = TRUE), 1), " - ", round(max(x, na.rm = TRUE), 1))
}

# Create the "COVID-19 Total" subset
covid_total <- df_clinical %>%
  filter(`Group d` %in% c(2, 3)) %>%
  mutate(Group_Label = "COVID-19 Total")

# Bind them together and set factor levels for correct column order later
combined_clinical <- bind_rows(df_clinical, covid_total) %>%
  mutate(Group_Label = factor(Group_Label, levels = c(
    "Healthy Control", "Non-COVID-19", "COVID-19 Total", "Non-Severe COVID-19", "Severe COVID-19"
  )))

# Calculate all summary statistics in one elegant pipeline
summary_stats <- combined_clinical %>%
  group_by(Group_Label) %>%
  summarise(
    # Counts
    N = n(),
    n_male = sum(`Sex g` == 1, na.rm = TRUE),
    n_female = sum(`Sex g` == 0, na.rm = TRUE),
    per_male = (n_male / N) * 100,
    per_female = (n_female / N) * 100,

    # Age & BMI
    across(c(`Age (year)`, `BMI h`, days_onset_to_adm, days_adm_to_severe), 
           list(
             mean = ~mean(., na.rm = TRUE),
             sd = ~sd(., na.rm = TRUE),
             median = ~median(., na.rm = TRUE),
             iqr = ~IQR(., na.rm = TRUE),
             range = ~get_range(.)
           ), 
           .names = "{.col}_{.fn}")
  ) %>%
  # Round all numeric values to 2 decimal places
  mutate(across(where(is.numeric), ~ round(., 2)))

# Format and Render Table 1
# Format the strings 
formatted_table <- summary_stats %>%
  mutate(
    # Using sprintf ensures trailing zeros are kept (e.g., 12.50 instead of 12.5)
    Male = sprintf("%d (%.1f%%)", n_male, per_male),
    Female = sprintf("%d (%.1f%%)", n_female, per_female),
    
    `Age Mean ± SD`  = sprintf("%.2f ± %.2f", `Age (year)_mean`, `Age (year)_sd`),
    `Age Median(IQR)` = sprintf("%.2f ± %.2f", `Age (year)_median`, `Age (year)_iqr`),
    `Age Range` = `Age (year)_range`,
    
    `BMI Mean ± SD`  = sprintf("%.2f ± %.2f", `BMI h_mean`, `BMI h_sd`),
    `BMI Median(IQR)` = sprintf("%.2f ± %.2f", `BMI h_median`, `BMI h_iqr`),
    `BMI Range` = `BMI h_range`,
    
    `Onset Mean ± SD`  = sprintf("%.2f ± %.2f", days_onset_to_adm_mean, days_onset_to_adm_sd),
    `Onset Median(IQR)` = sprintf("%.2f ± %.2f", days_onset_to_adm_median, days_onset_to_adm_iqr),
    `Onset Range` = days_onset_to_adm_range,
    
    `Severe Mean ± SD`  = sprintf("%.2f ± %.2f", days_adm_to_severe_mean, days_adm_to_severe_sd),
    `Severe Median(IQR)` = sprintf("%.2f ± %.2f", days_adm_to_severe_median, days_adm_to_severe_iqr),
    `Severe Range` = days_adm_to_severe_range
  ) %>%
  # Select only the Group and the formatted text columns
  select(Group_Label, N, Male, Female, contains("Mean ± SD"), contains("Median(IQR)"), contains("Range")) %>%
  # Add the N to the column header string
  mutate(Group_Label = paste0(Group_Label, " (N = ", N, ")")) %>%
  select(-N)

# Pivot the table so Groups are columns
final_table <- formatted_table %>%
  pivot_longer(cols = -Group_Label, names_to = "Metric_Raw", values_to = "Value") %>%
  pivot_wider(names_from = Group_Label, values_from = Value) %>%
  # Clean up the metric names to look nice in the table (e.g., "Age Mean ± SD" -> "Mean ± SD")
  mutate(Variables = gsub("^(Age|BMI|Onset|Severe) ", "", Metric_Raw)) %>%
  select(Variables, everything(), -Metric_Raw)

# Clean up irrelevant biological data with "-"
# (Healthy/Non-Covid don't have COVID onset dates; Non-severe don't have Severe dates)
final_table <- final_table %>%
  mutate(
    # Clean Healthy and Non-COVID columns (columns 2 and 3) for rows 9 to 14
    across(c(2, 3), ~ ifelse(row_number() >= 9 & row_number() <= 14, "-", .)),
    # Clean COVID Total and Non-Severe (columns 4 and 5) for rows 12 to 14
    across(c(4, 5), ~ ifelse(row_number() >= 12 & row_number() <= 14, "-", .))
  )

# Render with kableExtra
final_table %>%
  kbl(booktabs = TRUE, 
      caption = "Tabla 1. Características Clínicas y de Laboratorio de los Pacientes") %>%
  kable_classic(full_width = FALSE, html_font = "Arial") %>%
  pack_rows("Sex - no. (%)", 1, 2) %>%
  pack_rows("Age - year", 3, 5) %>%
  pack_rows("BMI, kg/m²", 6, 8) %>%
  pack_rows("Time from Onset to Admission, Days", 9, 11) %>%
  pack_rows("Time from Admission to Severe, Days", 12, 14) %>%
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, bold = TRUE, width = "10cm") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```


## Exercise 1.2
Perform an exploratory data analysis (EDA) of the dataset. At a minimum, include:

- Descriptive statistics for key variables  
- Appropriate visualizations (e.g. distributions, group comparisons)  
- A brief written summary of notable patterns, anomalies, or data quality issues  

Descriptive statistics 
```{r}
# Ungroup, filter, and select columns
df_labs <- combined_clinical %>%
  ungroup() %>%
  filter(Group_Label != "Healthy Control") %>% 
  select(
    Group_Label,
    WBC = `WBC count, ×109/L`, 
    Lymphocyte = `Lymphocyte count, ×109/L`, 
    Monocyte = `Monocyte count, ×109/L`,
    Platelet = `Platelet count, ×109/L`,
    CRP = `CRP i, mg/L`,
    ALT = `ALT j,  U/L`,
    AST = `AST k, U/L`,
    GGT = `GGT l, U/L`,
    TBIL = `TBIL m, μmol/L`,
    DBIL = `DBIL n, μmol/L`,
    Creatinine = `Creatinine, μmol/L`,
    Glucose = `Glucose, mmol/L`
  )

# Calculate raw summary statistics
summary_labs <- df_labs %>%
  group_by(Group_Label) %>%
  summarise(
    N = n(),
    across(where(is.numeric), 
           list(
             mean = ~mean(., na.rm = TRUE),
             sd = ~sd(., na.rm = TRUE),
             median = ~median(., na.rm = TRUE),
             iqr = ~IQR(., na.rm = TRUE),
             range = ~get_range(.)
           ), 
           .names = "{.col}_{.fn}")
  )

# Pivot, Format, and Reshape
final_lab_table <- summary_labs %>%
  pivot_longer(
    cols = -c(Group_Label, N), 
    names_to = c("Parameter", ".value"), 
    names_sep = "_"
  ) %>%
  mutate(
    `Mean ± SD` = sprintf("%.2f ± %.2f", mean, sd),
    `Median (IQR)` = sprintf("%.2f (%.2f)", median, iqr),
    `Range` = range
  ) %>%
  select(Group_Label, N, Parameter, `Mean ± SD`, `Median (IQR)`, Range) %>%
  pivot_longer(cols = c(`Mean ± SD`, `Median (IQR)`, Range), names_to = "Metric", values_to = "Value") %>%
  mutate(
    Group_Label = paste0(Group_Label, "\n(N = ", N, ")")
  ) %>%
  select(-N) %>%
  pivot_wider(names_from = Group_Label, values_from = Value)

# Filter and Sort
param_order <- c("WBC", "Lymphocyte", "Monocyte", "Platelet", 
                 "CRP", "ALT", "AST", "GGT", 
                 "TBIL", "DBIL", "Creatinine", "Glucose")

final_lab_table <- final_lab_table %>%
  mutate(
    Parameter = factor(Parameter, levels = param_order),
    Metric = factor(Metric, levels = c("Mean ± SD", "Median (IQR)", "Range"))
  ) %>%
  filter(!is.na(Parameter)) %>%
  arrange(Parameter, Metric) %>%
  select(-Parameter)

# Render with kableExtra
final_lab_table %>%
  kbl(booktabs = TRUE, 
      align = c("l", rep("c", 4)), 
      caption = "Table 2. Laboratory and Biochemical Parameters of the Patients") %>%
  kable_classic(full_width = FALSE, html_font = "Arial") %>%
  
  pack_rows("WBC count, ×109/L", 1, 3) %>%
  pack_rows("Lymphocyte count, ×109/L", 4, 6) %>%
  pack_rows("Monocyte count, ×109/L", 7, 9) %>%
  pack_rows("Platelet count, ×109/L", 10, 12) %>%
  pack_rows("CRP, mg/L", 13, 15) %>%
  pack_rows("ALT, U/L", 16, 18) %>%
  pack_rows("AST, U/L", 19, 21) %>%
  pack_rows("GGT, U/L", 22, 24) %>%
  pack_rows("TBIL, μmol/L", 25, 27) %>%
  pack_rows("DBIL, μmol/L", 28, 30) %>%
  pack_rows("Creatinine, μmol/L", 31, 33) %>%
  pack_rows("Glucose, mmol/L", 34, 36) %>%
  
  row_spec(0, bold = TRUE) %>% 
  column_spec(1, bold = FALSE, width = "4cm") %>%
  kable_styling(latex_options = c("hold_position", "scale_down"))
```
Study of data distribution
```{r}
# Exploratory Data Analysis of Clinical Parameters
# Data Preparation
param_order <- c("WBC", "Lymphocyte", "Monocyte", "Platelet", 
                 "CRP", "ALT", "AST", "GGT", 
                 "TBIL", "DBIL", "Creatinine", "Glucose")

# Prepare the dataframe for plotting
df_plot_covid <- df_clinical %>%
  # Combine groups 2 and 3 to create group 5 (Total)
  bind_rows(df_clinical %>% filter(`Group d` %in% 2:3) %>% mutate(`Group d` = 5)) %>%
  filter(`Group d` %in% c(2, 3, 5)) %>%
  mutate(
    Group_Label = case_when(
      `Group d` == 2 ~ "Non-Severe",
      `Group d` == 3 ~ "Severe",
      `Group d` == 5 ~ "Total COVID"
    ),
    # Order the groups: Total first, then Non-Severe, then Severe
    Group_Label = factor(Group_Label, levels = c("Total COVID", "Non-Severe", "Severe"))
  ) %>%
  # Rename columns for cleaner plot labels
  select(Group_Label, 
         WBC = `WBC count, ×109/L`, 
         Lymphocyte = `Lymphocyte count, ×109/L`, 
         Monocyte = `Monocyte count, ×109/L`,
         Platelet = `Platelet count, ×109/L`,
         CRP = `CRP i, mg/L`,
         ALT = `ALT j,  U/L`,
         AST = `AST k, U/L`,
         GGT = `GGT l, U/L`,
         TBIL = `TBIL m, μmol/L`,
         DBIL = `DBIL n, μmol/L`,
         Creatinine = `Creatinine, μmol/L`,
         Glucose = `Glucose, mmol/L`) %>%
  pivot_longer(cols = -Group_Label, names_to = "Parameter", values_to = "Value") %>%
  mutate(Parameter = factor(Parameter, levels = param_order))

# Define Parameter Groups
vars_part1 <- c("WBC", "Lymphocyte", "Monocyte", "Platelet")
vars_part2 <- c("CRP", "ALT", "AST", "GGT")
vars_part3 <- c("TBIL", "DBIL", "Creatinine", "Glucose")

# Plotting Function
create_violin_plot <- function(data, param_list, plot_title) {
  
  subset_data <- data %>%
    filter(Parameter %in% param_list) %>%
    mutate(Parameter = factor(Parameter, levels = param_list))
  
  ggplot(subset_data, aes(x = Group_Label, y = Value, fill = Group_Label)) +
    geom_violin(trim = FALSE, alpha = 0.3, color = "grey60", linewidth = 0.3) +
    geom_jitter(aes(color = Group_Label), width = 0.2, size = 1.2, alpha = 0.5) +
    geom_boxplot(width = 0.15, outlier.colour = "red", outlier.shape = 17, 
                 outlier.size = 3, alpha = 0.6) +
    facet_wrap(~ Parameter, scales = "free_y", ncol = 2) +
    # Using the custom color palette established in previous exercises
    scale_fill_manual(values = c("Total COVID" = "#4169E1", "Non-Severe" = "#F4A460", "Severe" = "#8B4513")) +
    scale_color_manual(values = c("Total COVID" = "#4169E1", "Non-Severe" = "#F4A460", "Severe" = "#8B4513")) +
    theme_bw() +
    labs(title = plot_title, x = NULL, y = "Value") +
    theme(
      legend.position = "top",
      strip.text = element_text(face = "bold", size = 14), 
      axis.text.x = element_text(size = 11, face = "bold"),
      axis.title.y = element_text(size = 12, face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold", margin = ggplot2::margin(b = 15))
    )
}

# Generate Plots
# Simply calling the object renders it in the RMarkdown output; no need for print()
p1 <- create_violin_plot(df_plot_covid, vars_part1, "Hematology Parameters")
p2 <- create_violin_plot(df_plot_covid, vars_part2, "Inflammation & Liver Enzymes")
p3 <- create_violin_plot(df_plot_covid, vars_part3, "Bilirubin, Kidney & Glucose")

p1
p2
p3

```

Studying data distribution
```{r}
# Plotting Function for Density Plots
create_density_plot <- function(data, param_list, plot_title) {
  
  # Filter and order the data
  subset_data <- data %>%
    filter(Parameter %in% param_list) %>%
    mutate(Parameter = factor(Parameter, levels = param_list))
  
  # Build the density plot
  ggplot(subset_data, aes(x = Value, fill = Group_Label, color = Group_Label)) +
    geom_density(alpha = 0.4, linewidth = 0.8) +
    facet_wrap(~ Parameter, scales = "free", ncol = 2) +
    scale_fill_manual(values = c("Total COVID" = "#4169E1", "Non-Severe" = "#F4A460", "Severe" = "#8B4513")) +
    scale_color_manual(values = c("Total COVID" = "#4169E1", "Non-Severe" = "#F4A460", "Severe" = "#8B4513")) +
    
    # Formatting and aesthetics
    theme_bw() +
    labs(
      title = plot_title, 
      x = "Clinical Value", 
      y = "Density"
    ) +
    theme(
      legend.position = "top",
      legend.title = element_blank(),
      strip.text = element_text(face = "bold", size = 14), 
      axis.text = element_text(size = 11),
      axis.title = element_text(size = 12, face = "bold"),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 16, margin = ggplot2::margin(b = 15)),
      strip.background = element_rect(fill = "#f0f0f0", color = "black")
    )
}

# Generate the Density Plots
p_dens1 <- create_density_plot(df_plot_covid, vars_part1, "Density: Hematology Parameters")
p_dens2 <- create_density_plot(df_plot_covid, vars_part2, "Density: Inflammation & Liver Enzymes")
p_dens3 <- create_density_plot(df_plot_covid, vars_part3, "Density: Bilirubin, Kidney & Glucose")

p_dens1
p_dens2
p_dens3

```

Correlation study
```{r}
# Prepare the numeric data
df_corr <- combined_clinical %>%
  filter(Group_Label != "COVID-19 Total") %>%
  select(WBC = `WBC count, ×109/L`, Lym = `Lymphocyte count, ×109/L`, 
         Mon = `Monocyte count, ×109/L`, Plt = `Platelet count, ×109/L`, 
         CRP = `CRP i, mg/L`, ALT = `ALT j,  U/L`, AST = `AST k, U/L`, 
         GGT = `GGT l, U/L`, TBIL = `TBIL m, μmol/L`, DBIL = `DBIL n, μmol/L`, 
         Cre = `Creatinine, μmol/L`, Glu = `Glucose, mmol/L`) %>%
  mutate(across(everything(), as.numeric))

# Calculate Spearman correlation and P-values
cor_matrix <- cor(df_corr, use = "pairwise.complete.obs", method = "spearman")
res_p <- suppressWarnings(cor.mtest(df_corr, conf.level = 0.95, method = "spearman"))
p_matrix <- res_p$p

# Plotting only significance levels
# Adjusting 'mar' to ensure labels aren't cut off
par(mar = c(1, 1, 1, 1))

corrplot(cor_matrix, 
         method = "shade",            # Use colored squares for the background
         type = "upper",              # Upper triangle only
         order = "hclust",            # Cluster similar variables
         
         # --- SIGNIFICANCE (STARS) CONTROL ---
         p.mat = p_matrix, 
         insig = "label_sig",         # This puts stars (*, **, ***) in the boxes
         sig.level = c(0.001, 0.01, 0.05), # Thresholds for ***, **, *
         pch.cex = 1.5,               # SIZE of the stars (increase to make them pop)
         pch.col = "black",           # Color of the stars
         
         # --- AXIS & LABEL CONTROL ---
         tl.col = "black",            # Label color
         tl.srt = 45,                 # Label rotation
         tl.cex = 1.1,                # SIZE of axis labels

         # --- NUMBERS CONTROL ---
         addCoef.col = NULL,          
         
         diag = FALSE,                # Hide self-correlation
         col = COL2('BrBG', 10)       # Red-Blue diverging palette
)
```

# Exercise 2
## Exercise 2.1
Reproduce Supplementary Figure 1 using the data without modifying or removing any observations. Use the same variables and groupings as in the original figure. 
```{r}
# Filter for just the 3 groups we want to compare and drop empty factor levels
df_clean <- combined_clinical %>%
  filter(Group_Label %in% c("Non-COVID-19", "Non-Severe COVID-19", "Severe COVID-19")) %>%
  mutate(Group_Label = droplevels(Group_Label))

vars_of_interest <- c(
  "Age (year)", "BMI h", "WBC count, ×109/L", "Lymphocyte count, ×109/L", 
  "Monocyte count, ×109/L", "Platelet count, ×109/L", "CRP i, mg/L", 
  "ALT j,  U/L", "AST k, U/L", "GGT l, U/L", "TBIL m, μmol/L", 
  "DBIL n, μmol/L", "Creatinine, μmol/L", "Glucose, mmol/L"
)

results_table <- map_dfr(vars_of_interest, function(var) {
  df_clean %>%
    select(Group_Label, Value = all_of(var)) %>%
    drop_na() %>%
    # Run pairwise Welch tests with BH adjustment based on the new Group_Label
    t_test(Value ~ Group_Label, var.equal = FALSE, p.adjust.method = "BH") %>%
    mutate(
      comparison = paste0("p.adj_", group1, "_vs_", group2),
      p_formatted = ifelse(p.adj < 0.001, "< 0.001", as.character(round(p.adj, 4)))
    ) %>%
    select(comparison, p_formatted) %>%
    pivot_wider(names_from = comparison, values_from = p_formatted) %>%
    mutate(Variable = var, .before = 1)
})

vars_12 <- c(
  "WBC count, ×109/L", "Lymphocyte count, ×109/L", "Monocyte count, ×109/L",
  "Platelet count, ×109/L", "CRP i, mg/L", "ALT j,  U/L",
  "AST k, U/L", "GGT l, U/L", "TBIL m, μmol/L", 
  "DBIL n, μmol/L", "Creatinine, μmol/L", "Glucose, mmol/L"
)

titles_12 <- c(
  "WBC", "Lymphocyte", "Monocyte", "Platelet", "CRP", "ALT",
  "AST", "GGT", "TBIL", "DBIL", "Creatinine", "Glucose"
)

# Set up the exact comparisons based on your new factor labels
my_comparisons <- list(
  c("Non-COVID-19", "Non-Severe COVID-19"),
  c("Non-Severe COVID-19", "Severe COVID-19"),
  c("Non-COVID-19", "Severe COVID-19")
)

plot_list <- purrr::map2(vars_12, titles_12, function(var_name, plot_title) {

  temp_df <- df_clean %>%
    select(Group_Label, Value = all_of(var_name)) %>%
    drop_na()

  # Calculate Welch Stats with BH adjustment for the plot brackets
  temp_stats <- temp_df %>%
    t_test(Value ~ Group_Label,
           comparisons = my_comparisons,
           var.equal = FALSE,
           p.adjust.method = "BH") %>%
    add_significance(p.col = "p.adj") %>%
    add_xy_position(x = "Group_Label") %>%
    filter(p.adj < 0.05) # Keep only significant adjusted brackets

  # Create Plot
  ggplot(temp_df, aes(x = Group_Label, y = Value, color = Group_Label)) +
    geom_bar(stat = "summary", fun = "mean", fill = "white", 
             linewidth = 1, width = 0.6, show.legend = FALSE) +
    stat_summary(fun.data = mean_sd, geom = "errorbar", 
                 width = 0.2, linewidth = 0.7, show.legend = FALSE) +
    geom_jitter(aes(shape = Group_Label), 
                width = 0.2, height = 0, size = 1.2, alpha = 0.8) +
    stat_pvalue_manual(temp_stats, label = "p.adj.signif", 
                       tip.length = 0.01, size = 3.5, bracket.size = 0.6) +
    # Updated to match your exact new group labels
    scale_color_manual(values = c("Non-COVID-19" = "#4169E1", 
                                  "Non-Severe COVID-19" = "#F4A460", 
                                  "Severe COVID-19" = "#8B4513")) +
    scale_shape_manual(values = c("Non-COVID-19" = 16, 
                                  "Non-Severe COVID-19" = 15, 
                                  "Severe COVID-19" = 17)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
    theme_classic() +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
      axis.line = element_line(color = "black", linewidth = 0.6),
      axis.text.x = element_blank(), 
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(size = 8)
    )
})

# ==========================================
# PART 3: FINAL COMPOSITION
# ==========================================
final_composition <- ggarrange(
  plotlist = plot_list,
  ncol = 3, nrow = 4,
  common.legend = TRUE,
  legend = "right"
)

print(final_composition)
```


## Exercise 2.2
Identify and handle outliers using an appropriate and well-justified method (e.g. removal, transformation, or robust statistics).

Reproduce Supplementary Figure 1 after outlier handling and discuss how and why the results change.
```{r}
# Variables to clean (the 12 lab parameters)
vars_to_clean <- c(
  "WBC count, ×109/L", "Lymphocyte count, ×109/L", "Monocyte count, ×109/L",
  "Platelet count, ×109/L", "CRP i, mg/L", "ALT j,  U/L",
  "AST k, U/L", "GGT l, U/L", "TBIL m, μmol/L", 
  "DBIL n, μmol/L", "Creatinine, μmol/L", "Glucose, mmol/L"
)

# All variables for the table (includes Age and BMI)
vars_of_interest <- c("Age (year)", "BMI h", vars_to_clean)

# Define the function to detect outliers (1.5x IQR)
remove_outliers <- function(x) {
  q1 <- quantile(x, 0.25, na.rm = TRUE)
  q3 <- quantile(x, 0.75, na.rm = TRUE)
  iqr <- q3 - q1
  lower_bound <- q1 - 1.5 * iqr
  upper_bound <- q3 + 1.5 * iqr
  ifelse(x < lower_bound | x > upper_bound, NA, x)
}

# Prepare data
df_clean <- combined_clinical %>%
  # Keep only the 3 groups we want to plot and test
  filter(Group_Label %in% c("Non-COVID-19", "Non-Severe COVID-19", "Severe COVID-19")) %>%
  # Drop the empty factor levels (Healthy Control, COVID-19 Total)
  mutate(Group_Label = droplevels(Group_Label)) %>%
  # Group by our clean labels to calculate IQR strictly PER GROUP
  group_by(Group_Label) %>%
  # Apply outlier removal just to the 12 lab vars
  mutate(across(all_of(vars_to_clean), remove_outliers)) %>%
  ungroup()

# Check: See how many NAs (outliers) were created per variable
print("Outliers removed per variable:")
print(colSums(is.na(df_clean[vars_to_clean])))

# ==========================================
# compute Welch t-test using Benjamini-Hochberg correction method
results_table <- map_dfr(vars_of_interest, function(var) {
  df_clean %>%
    select(Group_Label, Value = all_of(var)) %>%
    drop_na() %>%
    t_test(Value ~ Group_Label, var.equal = FALSE, p.adjust.method = "BH") %>%
    mutate(
      comparison = paste0("p.adj_", group1, "_vs_", group2),
      p_formatted = ifelse(p.adj < 0.001, "< 0.001", as.character(round(p.adj, 4)))
    ) %>%
    select(comparison, p_formatted) %>%
    pivot_wider(names_from = comparison, values_from = p_formatted) %>%
    mutate(Variable = var, .before = 1)
})

titles_12 <- c(
  "WBC", "Lymphocyte", "Monocyte", "Platelet", "CRP", "ALT",
  "AST", "GGT", "TBIL", "DBIL", "Creatinine", "Glucose"
)

# Exact comparisons matching your preprocessed factor labels
my_comparisons <- list(
  c("Non-COVID-19", "Non-Severe COVID-19"),
  c("Non-Severe COVID-19", "Severe COVID-19"),
  c("Non-COVID-19", "Severe COVID-19")
)

plot_list <- purrr::map2(vars_to_clean, titles_12, function(var_name, plot_title) {

  temp_df <- df_clean %>%
    select(Group_Label, Value = all_of(var_name)) %>%
    drop_na()

  # Calculate Welch Stats with BH adjustment
  temp_stats <- temp_df %>%
    t_test(Value ~ Group_Label,
           comparisons = my_comparisons,
           var.equal = FALSE,
           p.adjust.method = "BH") %>%
    add_significance(p.col = "p.adj") %>%
    add_xy_position(x = "Group_Label") %>%
    filter(p.adj < 0.05) # Only include significant adjusted brackets

  # Create Plot
  ggplot(temp_df, aes(x = Group_Label, y = Value, color = Group_Label)) +
    geom_bar(stat = "summary", fun = "mean", fill = "white", 
             linewidth = 1, width = 0.6, show.legend = FALSE) +
    stat_summary(fun.data = mean_sd, geom = "errorbar", 
                 width = 0.2, linewidth = 0.7, show.legend = FALSE) +
    geom_jitter(aes(shape = Group_Label), 
                width = 0.2, height = 0, size = 1.2, alpha = 0.8) +
    stat_pvalue_manual(temp_stats, label = "p.adj.signif", 
                       tip.length = 0.01, size = 3.5, bracket.size = 0.6) +
    # Colors and shapes mapped to the labels
    scale_color_manual(values = c("Non-COVID-19" = "#4169E1", 
                                  "Non-Severe COVID-19" = "#F4A460", 
                                  "Severe COVID-19" = "#8B4513")) +
    scale_shape_manual(values = c("Non-COVID-19" = 16, 
                                  "Non-Severe COVID-19" = 15, 
                                  "Severe COVID-19" = 17)) +
    scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.18))) +
    theme_classic() +
    labs(title = plot_title, x = NULL, y = NULL) +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 11),
      axis.line = element_line(color = "black", linewidth = 0.6),
      axis.text.x = element_blank(), 
      axis.ticks.x = element_blank(),
      axis.text.y = element_text(size = 8)
    )
})

# Create the composition of plots
final_composition <- ggarrange(
  plotlist = plot_list,
  ncol = 3, nrow = 4,
  common.legend = TRUE,
  legend = "right"
)

print(final_composition)
```

# Exercise 3
Create a heatmap of the biomarker data. Include group and gender as annotation variables.

```{r}
# Define Variables and Titles
vars_12 <- c(
  "WBC count, ×109/L", "Lymphocyte count, ×109/L", "Monocyte count, ×109/L",
  "Platelet count, ×109/L", "CRP i, mg/L", "ALT j,  U/L",
  "AST k, U/L", "GGT l, U/L", "TBIL m, μmol/L", 
  "DBIL n, μmol/L", "Creatinine, μmol/L", "Glucose, mmol/L"
)

titles_12 <- c(
  "WBC", "Lymphocyte", "Monocyte",
  "Platelet", "CRP", "ALT",
  "AST", "GGT", "TBIL", 
  "DBIL", "Creatinine", "Glucose"
)

# Filter and Prepare Data
df_plot <- combined_clinical %>%
  # Filter out Healthy Controls and the COVID-19 Total aggregate group
  filter(Group_Label %in% c("Non-COVID-19", "Non-Severe COVID-19", "Severe COVID-19")) %>%
  mutate(
    # Ensure Group is a clean factor
    Group_Label_Clean = droplevels(factor(Group_Label)),
    
    # Handle Sex: Map 1 to Male, 0 to Female (Matching Table 1 logic)
    Sex_Clean = factor(`Sex g`, 
                       levels = c(1, 0), 
                       labels = c("Male", "Female"))
  )

# Prepare the Matrix
# Extract biomarker columns and convert to numeric matrix
mat_data <- df_plot %>% select(all_of(vars_12))
colnames(mat_data) <- titles_12
mat <- as.matrix(mat_data)

# Assign Patient Names to rows 
rownames(mat) <- df_plot$`Patient ID a`  

# Prepare Annotation DataFrame
# ---------------------------
annotation_df <- data.frame(
  Group = df_plot$Group_Label_Clean,
  Sex   = df_plot$Sex_Clean
)
# Ensure row names match the matrix exactly
rownames(annotation_df) <- rownames(mat)

# Clean Missing Values (Prevent NA/NaN/Inf Error in clustering)
valid_rows <- complete.cases(mat)
mat <- mat[valid_rows, ]
annotation_df <- annotation_df[valid_rows, , drop = FALSE]

# Define Custom Colors
ann_colors <- list(
  Group = c(
    "Non-COVID-19"        = "#4169E1", 
    "Non-Severe COVID-19" = "#F4A460", 
    "Severe COVID-19"     = "#8B4513"
  ),
  Sex = c(
    "Male"   = "#0ead69", 
    "Female" = "#ee4266"
  )
)

# Generate Heatmap
pheatmap(mat,
         annotation_row = annotation_df,
         annotation_colors = ann_colors,
         color = colorRampPalette(c("blue", "white", "red"))(50),
         scale = "column",          
         main = "Biomarker Levels by COVID Severity",
         
         # --- THE "SPACE" & MARGIN FIXES ---
         show_rownames = TRUE,      
         fontsize_row = 3,          
         cellheight = 3,            
         cellwidth = 20,            
         
         # Increase these to push the heatmap away from the edges
         treeheight_row = 50,       # Pushes heatmap to the right
         treeheight_col = 35,       # Pushes heatmap down
         
         # This adds specific padding (in points) for the bottom/right/top/left
         # Usually, adding space to the right (margin 4) helps with row names
         # pheatmap doesn't have a direct 'mar' argument, so we use this:
         clustering_distance_rows = "euclidean", 
         
         # --- AESTHETICS ---
         border_color = "grey90",      
         cluster_rows = TRUE,
         cluster_cols = TRUE,
         angle_col = 45,
         fontsize_col = 10,
         
         # Ensure the legend doesn't overlap the plot
         annotation_legend = TRUE,
         legend = TRUE
)
```

# Conclusion

Based on the analysis made and the data shown, there is a significant *divergence* between **healthy** controls, **non-severe COVID-19** patients and  **severe COVID-19** patients. Some of the patterns observed depict different clinical profiles across the different groups: Patients in the *severe* group, have an above average **BMI** and **Age** than other groups with biomarkers that provide evidence on systemic inflammation and organ dysfunction; other features that determine this group is a significantly higher levels of **CRP** and **WBC** counts, alongside a decrease in **Lymphocytes** (severe viral infection). There is also indicators of liver and kidney stress with markers as **ALT**, **AST** and **Creatinine** being more elevated in the severe group, suggesting that the severity of COVID-19 is linked to multiple organs and a hyper-inflammatory response.

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```